\thesistitle

### 2019-08-11 sections/clonerefactor.tex
Proper capitalization for some concepts

Figure \ref{fig:clonerefactorprocess} shows the overall process used by CloneRefactor. First, we detect clones on the basis of a Java codebase, given a Java project from disk and a configuration. The clone detection process is further explained in section \ref{sec:clonedetection}. After all clones are found, CloneRefactor maps the context of the clones. On the basis of this context, CloneRefactor applies transformations to the source code for clones for which we have configured a refactoring.

@@

In order to be able to trace referenced identifiers SymbolSolver requires access to not only the analyzed Java projects, but also all its dependencies. This requires us to include all dependencies with the project. Along with this, SymbolSolver solves symbols in the JRE System Library (the standard libraries coming with every installation of Java) using the active Java Virtual Machine (JVM). This has a big impact on performance efficiency. This is visible in our results, as displayed in section \ref{sec:clonetypeexperiments}.

@@

When we analyze type 2R or type 3R clones we use the following additional threshold:
\begin{itemize}
  \item \textbf{T2R Variability}: The percentage of variability we allow in variables, method calls and literals. How this threshold is calculated is explained in section \ref{sec:t2rcheckglobalthres}.
\end{itemize}

When we analyze type 3 or type 3R clones we use the following additional threshold:
\begin{itemize}
  \item \textbf{T3R Gap Size}: The size of the gap we allow between two valid clones. This is a percentage of the size of the gap against the size of both clones combined. How this threshold is calculated is further explained in section \ref{sec:t3rclonerefactor}.
\end{itemize}

@@

The overall process regarding clone detection is displayed in figure \ref{fig:clonedetection}. First of all, we use JavaParser to read a project from disk and build an AST, one class file at a time. Each AST is then converted to a directed graph that maps relations between statements, further explained in section \ref{sec:clonegraph}. On the basis of this graph, we detect clone classes and verify them using three thresholds (in order of importance):
\begin{itemize}
  \item Amount of tokens.
  \item Amount of statements.
  \item Amount of lines.
\end{itemize}
If the detection was configured to detect either type 2R, 3 or 3R clones we perform some type specific transformations on the resulting set of clones.

@@

\subsection{Generating the clone graph}\label{sec:clonegraph}
First of all, we parse the AST obtained from JavaParser into a directed graph structure. We have chosen to base our clone detection around statements as the smallest unit of comparison. This means that a single statement cloned with another single statement is the smallest clone we can find. The rationale for this lies in both simplicity and performance efficiency. This means we won't be able to find when a single expression matches another expression, or even a single token matching another token. This is in most cases not a problem, as expressions are often small and do not span the minimal size to be considered a clone in the first place (more about this in section \ref{sec:thresholds}).

@@

\subsubsection{Building the clone graph}\label{sec:buildingclonegraph}
Building the clone graph consists of walking the AST in-order for each declaration and statement. For each declaration/statement found, we map the following relations:
\begin{itemize}
  \item The declaration/statement preceding it.
  \item The declaration/statement following.
  \item The last \textbf{preceding} declaration/statement with which it is cloned.
\end{itemize}
We do not create a separate graph for each class file, so the statement/declaration preceding or following could be in a different file. While mapping these relations, we maintain a hashed map containing the last occurrence of each unique statement. This map is used to efficiently find out whether a statement is cloned with another. An example of such a graph is displayed in figure \ref{fig:clonegraphsimple}.

@@

We refer to the declarations and statements in this graph as \textit{nodes}. The relations \textit{next} and \textit{previous} in this graph are represented as an twodirectional arrow. The relations representing duplication are directed. This is a restriction we've chosen as it creates an important constraint for the clone detection process. This process is explained in section \ref{sec:detectingclones}.

@@

\subsubsection{Mapping graph nodes to code}
The clone graph, as explained in section \ref{sec:buildingclonegraph}, contains all declarations and statements in a source code. However, declarations and statements may themselves have child declarations and statements. To avoid redundant duplication checks, we exclude the body of each node. Look at figure \ref{fig:clonegraph} for an example of how source code maps to AST nodes.

@@

As stated in the previous section, nodes in the graph link to \textit{preceding} cloned statement. This implies that the first node that is cloned does not have any clone relation, as there are no clones preceding it (only following it). Because of this, we start our clone detection process at the final location encountered while building the graph. For this node, we collect all nodes it is cloned with. Even though the final node only links to the preceding node it is cloned with, we can collect all clones. This is because the preceding clone also has a preceding clone (if applicable) and we can follow this trail to collect all clones of a single node. As an example, we convert the code example shown in figure \ref{fig:clonegraph} to a clone graph as displayed in figure \ref{fig:clonegraph2}.

@@

Using the example shown in figure \ref{fig:clonegraph} and \ref{fig:clonegraph2} we can explain how we detect clones on the basis of this graph. Suppose we are finding clones for two files and the final node of the second file is a variable declarator. This final node is represented in the example figure by the purple box (1). We then follow all ``duplicate'' relations until we have found all clones of this node (2 and 3). We now have a single statement clone class of three clone instances (1, 2 and 3).

@@

\subsection{Validating the type 2R variability threshold}
In the definition of type 2R clones (see section \ref{sec:type2r}) we described how type 2R clones work on the basis of a variability threshold. This threshold is checked by CloneRefactor to assess the validity of found clone classes. Implementing such a threshold involves some important design decisions and has a lot of complexity. In this section we explain how CloneRefactor detects type 2R clones on the basis of this variability threshold. This process is done as a postprocessing step after clone detection. The type 2R clone detection process is described in more detail in section \ref{sec:comparingstuff}. In short, this process detects clones allowing for any variability between expressions. This postprocessing step then determines which (parts of) these clones are valid by the configure variability threshold.

@@

\subsubsection{Create equality array}
To determine the difference in literals, method calls and variables, we convert the code to an equality array. This equality array converts each (group of) token(s) to a number unique to that (group of) tokens. Each literal, method call or variable becomes a positive number, whereas each other token becomes a negative number. An example of two code fragments converted to equality arrays is displayed in figure \ref{fig:equalityarrays}.

@@

The equality array for each of the lines in this example is shown in table \ref{table:equalityarrays}.

\begin{table}[H]
\begin{center}
 \caption{The equality arrays in our example figure \ref{fig:equalityarrays}.} \label{table:equalityarrays}
 \medskip
\begin{tabular}{|l|l|l|}
\hline
Node (n) & Equality Array (E) \\ \hline
1        & {[}1, -2, 3, -5, -6{]} \\ \hline
2        & {[}1, -2, 7, -5, -6{]} \\ \hline
3        & {[}1, -2, 4, -5, -6{]} \\ \hline
4        & {[}8, -2, 9, -5, -6{]} \\ \hline
\end{tabular}
\end{center}
\end{table}

In the example of figure \ref{fig:equalityarrays} the fragment on the left (consisting of $n_1$ and $n_2$) and the fragment on the right (consisting of $n_3$ and $n_4$) are two clone instances of a clone class. Table \ref{table:equalityarrays} shows their corresponding equality arrays.

\subsubsection{Checking global thresholds} \label{sec:t2rcheckglobalthres}
Using the equality arrays explained in the previous section, we can determine the variability threshold of any clone class. We calculate the variability of the example given in table \ref{table:equalityarrays} as follows:

@@

Where $E_x$ refers to the equality arrays shown in table \ref{table:equalityarrays} and \textit{x} is the node number displayed in the same table. When we apply this equation to the example clone classes displayed in figure \ref{fig:equalityarrays}, we get the following sum:

@@

\todo{Usage of the word adheres is questionable here...}
So for the example given in figure \ref{fig:equalityarrays} we have a variability of 75\%. In CloneRefactor, the maximum variability percentage is a threshold that is entered in a configuration file. If a clone adheres to this threshold, it will stay in the set of found clones. However, if it does not adhere to the thresholds, a problem arises. Because a clone does not adhere to the thresholds, it does not yet mean it has to be discarded. This is because an invalid clone class can still contain valid clones that are a subset of the invalid clone (our definition of subsets of clones is given in section \ref{sec:conceptualremovingredundant}).

@@

The first line of the cloned fragment shown in figure \ref{fig:2rvariabilityhigh1} has a high variability with its cloned fragement. However, the rest of this method does not have any variability. The global thresholds could indicate a too high variability and thus render this clone invalid. However, in this case, it might still have a valid subclone.

@@

Another example of high variability between clones, in which a valid subclone can be found, is displayed in figure \ref{fig:2rvariabilityhigh2}. In this case, one clone instance has such a high variability that it shouldn't be refactored. In this case, the clone instance with high variability should be removed.

@@

\begin{figure}[H]
\begin{parcolumns}{3}
\colchunk[1]{
\begin{javacode}
|\highlightYellow|doA(); //{1, -2}
|\highlightYellow|doB(); //{3, -2}
doC(); //{4, -2}
doC(); //{4, -2}
\end{javacode}}
\colchunk[2]{
\begin{javacode}
doD(); //{5, -2}
|\highlightYellow|doA(); //{1, -2}
|\highlightYellow|doB(); //{3, -2}
doD(); //{5, -2}
\end{javacode}}
\colchunk[3]{
\begin{javacode}
doE(); //{6, -2}
doE(); //{6, -2}
|\highlightYellow|doA(); //{1, -2}
|\highlightYellow|doB(); //{3, -2}
\end{javacode}}
\end{parcolumns}
\caption{Cloned code fragments from figure \ref{fig:2rvariabilityhigh3} together with their equality arrays.}
\label{fig:2rvariabilityhigh4}
\end{figure}

In figure \ref{fig:2rvariabilityhigh4} we display clone classes and their (simplified) equality arrays. We can convert this to a graph, using a similar process as the graph creation process used for clone detection. In this case, duplicate relations represent nodes of which their equality arrays are within the variability threshold. In figure \ref{fig:2rvariabilityhigh4} all relations are between equal equality arrays, but this does not always have to be the case. Large equality arrays, denoting statements that consist of many tokens, may have some variability and still be considered duplicates in this graph.

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{img/T2RGraph}
  \caption{Graph representation of the code example displayed in figure \ref{fig:2rvariabilityhigh4}.}
  \label{fig:clonerefactorprocess}
\end{figure}

@@

\subsubsection{Detect Clones on Graph}
Similar to the clone detection process, we detect clones on the basis of the graph described in the previous section. The only difference is that, in this step, we do not remove clones that do not meet the thresholds (as explained in section \ref{sec:clonerefactorthresholds}). This is because the next step, explained in the next section, could potentially expand the found clones and thus make clones that would currently invalid by the thresholds valid.

Looking back at the example of the previous section, it would result in the following clone class collection:
\begin{figure}[H]
\begin{parcolumns}{3}
\colchunk[1]{
\begin{javacode}
|\highlightYellow|doA(); // |\textcolor{pgreen}{$C_1$ $I_3$}|
|\highlightYellow|doB(); // |\textcolor{pgreen}{$C_1$ $I_3$}|
doC(); // |\textcolor{pgreen}{$C_4$ $I_2$}|
doC(); // |\textcolor{pgreen}{$C_4$ $I_1$}|
\end{javacode}}
\colchunk[2]{
\begin{javacode}
doD(); // |\textcolor{pgreen}{$C_3$ $I_2$}|
|\highlightYellow|doA(); // |\textcolor{pgreen}{$C_1$ $I_2$}|
|\highlightYellow|doB(); // |\textcolor{pgreen}{$C_1$ $I_2$}|
doD(); // |\textcolor{pgreen}{$C_3$ $I_1$}|
\end{javacode}}
\colchunk[3]{
\begin{javacode}
doE(); // |\textcolor{pgreen}{$C_2$ $I_2$}|
doE(); // |\textcolor{pgreen}{$C_2$ $I_1$}|
|\highlightYellow|doA(); // |\textcolor{pgreen}{$C_1$ $I_1$}|
|\highlightYellow|doB(); // |\textcolor{pgreen}{$C_1$ $I_1$}|
\end{javacode}}
\end{parcolumns}
\caption{Clone instances in figure \ref{fig:2rvariabilityhigh4} as determined .}
\label{fig:2rvariabilityhigh5}
\end{figure}

Where \textit{C} denotes the clone class and \textit{I} denotes the clone instance. The process by which these clones are found is equal to the clone detection process as described in section \ref{sec:detectingclones}. In this example, we see the four clones found on the graph of figure \ref{fig:clonerefactorprocess}. Three of these clone classes have a two clone instances, each consisting of a single nodes. The other clone class has three clone instances, each consisting of two nodes.

@@

As an example, consider the clone fragment in figure \ref{fig:trytoexpand}. In this example, we have 2 clone classes. In the ``Try To Expand'' step, we will check whether we can create a clone class with larger clone instances on the basis of the clones found in the previous step. We start with the largest clone class, measured in clone volume. We measure clone volume as the product of the amount of clone instances in a clone class and the amount of nodes in any of its instances.

In the example in figure \ref{fig:trytoexpand} the largest instance has a volume of:

@@

Starting from this clone, we will try to expand it. We will include the previous node in the clone class and verify its variability threshold (node $N_2$). In this case, using the formulas shown in section \ref{sec:t2rcheckglobalthres}, we can check whether this clone class would be valid by the variability threshold:

@@

This does not fall within the variability threshold, so we do not include this node in the clone class. After we have checked a single clone class within the bigger cloned fragment for expansion opportunities, we go on with the next clone class (by volume). In this case, the other clone class within this fragment has become a subset of the clone class we just expanded. Because of that, this other clone class is discarded. For the example code fragment in figure \ref{fig:trytoexpand} the resulting clone class is $\{N_1, N_2, N_3, N_4, N_5\}$.

\subsection{Checking for type 3 opportunities} \label{sec:t3rclonerefactor}
As described in section \ref{sec:type3r}, we define type 3R clones as gapped clones. This means that two existing clones may have a gap of non-gapped clones. We check for every found clone class whether it is a type 3R clone with another clone:

@@

In the example of figure \ref{fig:3ropportunities} we see two clone classes, which are separated by a single node. Using equation \ref{eq:t3rgap} we can calculate the size of the gap as follows:

@@

\subsubsection{Partial Method/Constructor}
These categories describe clone instances which are found in the body of a method or constructor. These clones can often be refactored by extracting a new method out of the cloned code. We detect such clones by checking that each node in the clone instance has a ancestor node that is a MethodDeclaration, but none of the nodes is a method declaration. We describe the procedure by with we recursively seek the parent nodes of a node in section \ref{sec:methodlevelcr}.

@@

Because of this, we defined a set of categories of things that can obstruct a refactoring opportunity using the ``Extract Method'' technique. We defined the following categories:
\begin{itemize}
    \item \textbf{Complex Control Flow}: This clone contains \texttt{break}, \texttt{continue} or \texttt{return} statements, obstructing the possibility of method extraction.
    \item \textbf{Spans Part Of A Block}: This clone spans a part of a statement.
    \item \textbf{Is Not A Partial Method}: If the clone does not fall in the ``Partial method'' category of section \ref{sec:setupcontents}, the ``extract method'' refactoring technique cannot be applied.
    \item \textbf{Top-level Node Is Not A Statement}: The topmost node of a clone instance is not a statement.
    \item \textbf{Overlaps}: There is overlap within the clone class.
    \item \textbf{Can Be Extracted}: This clone is a fragment of code that can directly be extracted to a new method. Then, based on the relation between the clone instances, further refactoring techniques can be used to refactor the extracted methods (for instance ``pull up method'' for clones in sibling classes).
\end{itemize}

This does not mean that clones outside the ``Can be extracted'' category cannot be refactored using method extraction. However, they may require additional transformations, other refactoring techniques, etc. Because of that, we took those categories out of the scope for our automated refactoring efforts. We further explain opportunities to refactor such clones outside the ``Can be extracted'' category in section \ref{sec:discussrefactorability}.

@@

\subsubsection{Spans Part Of A Block}
When applying an ``Abstract Method'' refactoring, we author a new method declaration. The body of this method declaration consists of a set of statements. However, if a single statement is only partially cloned, this might give issues with refactoring. An example of this is shown in figure \ref{fig:refactorabilitypartialblock}

@@

\subsubsection{Is Not A Partial Method}
The ``Extract Method'' refactoring can only refactor code in the body of a method. We mapped the contents of clones in section \ref{sec:setuplocation}. Here we defined the ``Partial Method'' catagory as \textit{a clone that spans (a part of) the body of a method. The declaration itself is not included.} If the clone does not fall within this category, ``Extract Method'' is likely not to be correct technique for refactoring such clones.

\subsubsection{Top-level Node Is Not A Statement}
A method body can only consist of statements. We cannot put declarations in a method. However, there are possibilities where a clone consists of something else than a statement as top level node. The most prominent example is when a clause is cloned (see section \ref{sec:astbackground} for examples of such concepts). For instance, we cannot extract only the catch-clause of a try-statement to a new method. In such a case, we should either extract the complete try-statement, or only the contents of the catch clause.

\subsubsection{Overlaps}
We do not consider clone classes for refactoring that overlap in the clone class itself. That is because such clones may require other techniques to refactor them. We consider analyzing such refactoring methods future work. An example refactoring of clone instances that overlap within their clone class is displayed in figure \ref{fig:overlaprefactoring}. In this example one clone instance is found between line \ref{line:overlaprefactoring1} and \ref{line:overlaprefactoring4} and the other between \ref{line:overlaprefactoring2} and \ref{line:overlaprefactoring5}. To refactor such clones, other techniques may be required, like introducing a new for-loop to reduce the duplication.

@@

\subsubsection{Relation}
For each of the identified clone relations (section \ref{sec:setuprelation}) we define a method of automatically refactoring clones in these categories. The relation determines the location at which we place the newly created method.

>>>>

Figure \ref{fig:clonerefactorprocess} shows the overall process used by CloneRefactor. First, we detect clones on the basis of a Java codebase, given a Java project from disk and a configuration. The clone detection process is further explained in Section~\ref{sec:clonedetection}. After all clones are found, CloneRefactor maps the context of the clones. On the basis of this context, CloneRefactor applies transformations to the source code for clones for which we have configured a refactoring.

@@

In order to be able to trace referenced identifiers SymbolSolver requires access to not only the analyzed Java projects, but also all its dependencies. This requires us to include all dependencies with the project. Along with this, SymbolSolver solves symbols in the JRE System Library (the standard libraries coming with every installation of Java) using the active Java Virtual Machine (JVM). This has a big impact on performance efficiency. This is visible in our results, as displayed in Section~\ref{sec:clonetypeexperiments}.

@@

When we analyze type 2R or type 3R clones we use the following additional threshold:
\begin{itemize}
  \item \textbf{T2R Variability}: The percentage of variability we allow in variables, method calls and literals. How this threshold is calculated is explained in Section~\ref{sec:t2rcheckglobalthres}.
\end{itemize}

When we analyze type 3 or type 3R clones we use the following additional threshold:
\begin{itemize}
  \item \textbf{T3R Gap Size}: The size of the gap we allow between two valid clones. This is a percentage of the size of the gap against the size of both clones combined. How this threshold is calculated is further explained in Section~\ref{sec:t3rclonerefactor}.
\end{itemize}

@@

The overall process regarding clone detection is displayed in Figure~\ref{fig:clonedetection}. First of all, we use JavaParser to read a project from disk and build an AST, one class file at a time. Each AST is then converted to a directed graph that maps relations between statements, further explained in Section~\ref{sec:clonegraph}. On the basis of this graph, we detect clone classes and verify them using three thresholds (in order of importance):
\begin{itemize}
  \item Amount of tokens.
  \item Amount of statements.
  \item Amount of lines.
\end{itemize}
If the detection was configured to detect either type 2R, 3 or 3R clones we perform some type specific transformations on the resulting set of clones.

@@

\subsection{Generating the clone graph}\label{sec:clonegraph}
First of all, we parse the AST obtained from JavaParser into a directed graph structure. We have chosen to base our clone detection around statements as the smallest unit of comparison. This means that a single statement cloned with another single statement is the smallest clone we can find. The rationale for this lies in both simplicity and performance efficiency. This means we won't be able to find when a single expression matches another expression, or even a single token matching another token. This is in most cases not a problem, as expressions are often small and do not span the minimal size to be considered a clone in the first place (more about this in Section~\ref{sec:thresholds}).

@@

\subsubsection{Building the clone graph}\label{sec:buildingclonegraph}
Building the clone graph consists of walking the AST in-order for each declaration and statement. For each declaration/statement found, we map the following relations:
\begin{itemize}
  \item The declaration/statement preceding it.
  \item The declaration/statement following.
  \item The last \textbf{preceding} declaration/statement with which it is cloned.
\end{itemize}
We do not create a separate graph for each class file, so the statement/declaration preceding or following could be in a different file. While mapping these relations, we maintain a hashed map containing the last occurrence of each unique statement. This map is used to efficiently find out whether a statement is cloned with another. An example of such a graph is displayed in Figure~\ref{fig:clonegraphsimple}.

@@

We refer to the declarations and statements in this graph as \textit{nodes}. The relations \textit{next} and \textit{previous} in this graph are represented as an twodirectional arrow. The relations representing duplication are directed. This is a restriction we've chosen as it creates an important constraint for the clone detection process. This process is explained in Section~\ref{sec:detectingclones}.

@@

\subsubsection{Mapping graph nodes to code}
The clone graph, as explained in Section~\ref{sec:buildingclonegraph}, contains all declarations and statements in a source code. However, declarations and statements may themselves have child declarations and statements. To avoid redundant duplication checks, we exclude the body of each node. Look at Figure~\ref{fig:clonegraph} for an example of how source code maps to AST nodes.

@@

As stated in the previous section, nodes in the graph link to \textit{preceding} cloned statement. This implies that the first node that is cloned does not have any clone relation, as there are no clones preceding it (only following it). Because of this, we start our clone detection process at the final location encountered while building the graph. For this node, we collect all nodes it is cloned with. Even though the final node only links to the preceding node it is cloned with, we can collect all clones. This is because the preceding clone also has a preceding clone (if applicable) and we can follow this trail to collect all clones of a single node. As an example, we convert the code example shown in Figure~\ref{fig:clonegraph} to a clone graph as displayed in Figure~\ref{fig:clonegraph2}.

@@

Using the example shown in Figure~\ref{fig:clonegraph} and \ref{fig:clonegraph2} we can explain how we detect clones on the basis of this graph. Suppose we are finding clones for two files and the final node of the second file is a variable declarator. This final node is represented in the example figure by the purple box (1). We then follow all ``duplicate'' relations until we have found all clones of this node (2 and 3). We now have a single statement clone class of three clone instances (1, 2 and 3).

@@

\subsection{Validating the type 2R variability threshold}
In the definition of type 2R clones (see Section~\ref{sec:type2r}) we described how type 2R clones work on the basis of a variability threshold. This threshold is checked by CloneRefactor to assess the validity of found clone classes. Implementing such a threshold involves some important design decisions and has a lot of complexity. In this section we explain how CloneRefactor detects type 2R clones on the basis of this variability threshold. This process is done as a postprocessing step after clone detection. The type 2R clone detection process is described in more detail in Section~\ref{sec:comparingstuff}. In short, this process detects clones allowing for any variability between expressions. This postprocessing step then determines which (parts of) these clones are valid by the configure variability threshold.

@@

\subsubsection{Create equality array}
To determine the difference in literals, method calls and variables, we convert the code to an equality array. This equality array converts each (group of) token(s) to a number unique to that (group of) tokens. Each literal, method call or variable becomes a positive number, whereas each other token becomes a negative number. An example of two code fragments converted to equality arrays is displayed in Figure~\ref{fig:equalityarrays}.

@@

The equality array for each of the lines in this example is shown in Table~\ref{table:equalityarrays}.

\begin{table}[H]
\begin{center}
 \caption{The equality arrays in our example Figure~\ref{fig:equalityarrays}.} \label{table:equalityarrays}
 \medskip
\begin{tabular}{|l|l|l|}
\hline
Node (n) & Equality Array (E) \\ \hline
1        & {[}1, -2, 3, -5, -6{]} \\ \hline
2        & {[}1, -2, 7, -5, -6{]} \\ \hline
3        & {[}1, -2, 4, -5, -6{]} \\ \hline
4        & {[}8, -2, 9, -5, -6{]} \\ \hline
\end{tabular}
\end{center}
\end{table}

In the example of Figure~\ref{fig:equalityarrays} the fragment on the left (consisting of $n_1$ and $n_2$) and the fragment on the right (consisting of $n_3$ and $n_4$) are two clone instances of a clone class. Table \ref{table:equalityarrays} shows their corresponding equality arrays.

\subsubsection{Checking global thresholds} \label{sec:t2rcheckglobalthres}
Using the equality arrays explained in the previous section, we can determine the variability threshold of any clone class. We calculate the variability of the example given in Table~\ref{table:equalityarrays} as follows:

@@

Where $E_x$ refers to the equality arrays shown in Table~\ref{table:equalityarrays} and \textit{x} is the node number displayed in the same table. When we apply this equation to the example clone classes displayed in Figure~\ref{fig:equalityarrays}, we get the following sum:

@@

\todo{Usage of the word adheres is questionable here...}
So for the example given in Figure~\ref{fig:equalityarrays} we have a variability of 75\%. In CloneRefactor, the maximum variability percentage is a threshold that is entered in a configuration file. If a clone adheres to this threshold, it will stay in the set of found clones. However, if it does not adhere to the thresholds, a problem arises. Because a clone does not adhere to the thresholds, it does not yet mean it has to be discarded. This is because an invalid clone class can still contain valid clones that are a subset of the invalid clone (our definition of subsets of clones is given in Section~\ref{sec:conceptualremovingredundant}).

@@

The first line of the cloned fragment shown in Figure~\ref{fig:2rvariabilityhigh1} has a high variability with its cloned fragement. However, the rest of this method does not have any variability. The global thresholds could indicate a too high variability and thus render this clone invalid. However, in this case, it might still have a valid subclone.

@@

Another example of high variability between clones, in which a valid subclone can be found, is displayed in Figure~\ref{fig:2rvariabilityhigh2}. In this case, one clone instance has such a high variability that it shouldn't be refactored. In this case, the clone instance with high variability should be removed.

@@

\begin{figure}[H]
\begin{parcolumns}{3}
\colchunk[1]{
\begin{javacode}
|\highlightYellow|doA(); //{1, -2}
|\highlightYellow|doB(); //{3, -2}
doC(); //{4, -2}
doC(); //{4, -2}
\end{javacode}}
\colchunk[2]{
\begin{javacode}
doD(); //{5, -2}
|\highlightYellow|doA(); //{1, -2}
|\highlightYellow|doB(); //{3, -2}
doD(); //{5, -2}
\end{javacode}}
\colchunk[3]{
\begin{javacode}
doE(); //{6, -2}
doE(); //{6, -2}
|\highlightYellow|doA(); //{1, -2}
|\highlightYellow|doB(); //{3, -2}
\end{javacode}}
\end{parcolumns}
\caption{Cloned code fragments from Figure~\ref{fig:2rvariabilityhigh3} together with their equality arrays.}
\label{fig:2rvariabilityhigh4}
\end{figure}

In Figure~\ref{fig:2rvariabilityhigh4} we display clone classes and their (simplified) equality arrays. We can convert this to a graph, using a similar process as the graph creation process used for clone detection. In this case, duplicate relations represent nodes of which their equality arrays are within the variability threshold. In Figure~\ref{fig:2rvariabilityhigh4} all relations are between equal equality arrays, but this does not always have to be the case. Large equality arrays, denoting statements that consist of many tokens, may have some variability and still be considered duplicates in this graph.

\begin{figure}[H]
  \centering
  \includegraphics[width=1\columnwidth]{img/T2RGraph}
  \caption{Graph representation of the code example displayed in Figure~\ref{fig:2rvariabilityhigh4}.}
  \label{fig:clonerefactorprocess}
\end{figure}

@@

\subsubsection{Detect Clones on Graph}
Similar to the clone detection process, we detect clones on the basis of the graph described in the previous section. The only difference is that, in this step, we do not remove clones that do not meet the thresholds (as explained in Section~\ref{sec:clonerefactorthresholds}). This is because the next step, explained in the next section, could potentially expand the found clones and thus make clones that would currently invalid by the thresholds valid.

Looking back at the example of the previous section, it would result in the following clone class collection:
\begin{figure}[H]
\begin{parcolumns}{3}
\colchunk[1]{
\begin{javacode}
|\highlightYellow|doA(); // |\textcolor{pgreen}{$C_1$ $I_3$}|
|\highlightYellow|doB(); // |\textcolor{pgreen}{$C_1$ $I_3$}|
doC(); // |\textcolor{pgreen}{$C_4$ $I_2$}|
doC(); // |\textcolor{pgreen}{$C_4$ $I_1$}|
\end{javacode}}
\colchunk[2]{
\begin{javacode}
doD(); // |\textcolor{pgreen}{$C_3$ $I_2$}|
|\highlightYellow|doA(); // |\textcolor{pgreen}{$C_1$ $I_2$}|
|\highlightYellow|doB(); // |\textcolor{pgreen}{$C_1$ $I_2$}|
doD(); // |\textcolor{pgreen}{$C_3$ $I_1$}|
\end{javacode}}
\colchunk[3]{
\begin{javacode}
doE(); // |\textcolor{pgreen}{$C_2$ $I_2$}|
doE(); // |\textcolor{pgreen}{$C_2$ $I_1$}|
|\highlightYellow|doA(); // |\textcolor{pgreen}{$C_1$ $I_1$}|
|\highlightYellow|doB(); // |\textcolor{pgreen}{$C_1$ $I_1$}|
\end{javacode}}
\end{parcolumns}
\caption{Clone instances in Figure~\ref{fig:2rvariabilityhigh4} as determined .}
\label{fig:2rvariabilityhigh5}
\end{figure}

Where \textit{C} denotes the clone class and \textit{I} denotes the clone instance. The process by which these clones are found is equal to the clone detection process as described in Section~\ref{sec:detectingclones}. In this example, we see the four clones found on the graph of Figure~\ref{fig:clonerefactorprocess}. Three of these clone classes have a two clone instances, each consisting of a single nodes. The other clone class has three clone instances, each consisting of two nodes.

@@

As an example, consider the clone fragment in Figure~\ref{fig:trytoexpand}. In this example, we have 2 clone classes. In the ``Try To Expand'' step, we will check whether we can create a clone class with larger clone instances on the basis of the clones found in the previous step. We start with the largest clone class, measured in clone volume. We measure clone volume as the product of the amount of clone instances in a clone class and the amount of nodes in any of its instances.

In the example in Figure~\ref{fig:trytoexpand} the largest instance has a volume of:

@@

Starting from this clone, we will try to expand it. We will include the previous node in the clone class and verify its variability threshold (node $N_2$). In this case, using the formulas shown in Section~\ref{sec:t2rcheckglobalthres}, we can check whether this clone class would be valid by the variability threshold:

@@

This does not fall within the variability threshold, so we do not include this node in the clone class. After we have checked a single clone class within the bigger cloned fragment for expansion opportunities, we go on with the next clone class (by volume). In this case, the other clone class within this fragment has become a subset of the clone class we just expanded. Because of that, this other clone class is discarded. For the example code fragment in Figure~\ref{fig:trytoexpand} the resulting clone class is $\{N_1, N_2, N_3, N_4, N_5\}$.

\subsection{Checking for type 3 opportunities} \label{sec:t3rclonerefactor}
As described in Section~\ref{sec:type3r}, we define type 3R clones as gapped clones. This means that two existing clones may have a gap of non-gapped clones. We check for every found clone class whether it is a type 3R clone with another clone:

@@

In the example of Figure~\ref{fig:3ropportunities} we see two clone classes, which are separated by a single node. Using equation \ref{eq:t3rgap} we can calculate the size of the gap as follows:

@@

\subsubsection{Partial Method/Constructor}
These categories describe clone instances which are found in the body of a method or constructor. These clones can often be refactored by extracting a new method out of the cloned code. We detect such clones by checking that each node in the clone instance has a ancestor node that is a MethodDeclaration, but none of the nodes is a method declaration. We describe the procedure by with we recursively seek the parent nodes of a node in Section~\ref{sec:methodlevelcr}.

@@

Because of this, we defined a set of categories of things that can obstruct a refactoring opportunity using the ``Extract Method'' technique. We defined the following categories:
\begin{itemize}
    \item \textbf{Complex Control Flow}: This clone contains \texttt{break}, \texttt{continue} or \texttt{return} statements, obstructing the possibility of method extraction.
    \item \textbf{Spans Part Of A Block}: This clone spans a part of a statement.
    \item \textbf{Is Not A Partial Method}: If the clone does not fall in the ``Partial method'' category of Section~\ref{sec:setupcontents}, the ``extract method'' refactoring technique cannot be applied.
    \item \textbf{Top-level Node Is Not A Statement}: The topmost node of a clone instance is not a statement.
    \item \textbf{Overlaps}: There is overlap within the clone class.
    \item \textbf{Can Be Extracted}: This clone is a fragment of code that can directly be extracted to a new method. Then, based on the relation between the clone instances, further refactoring techniques can be used to refactor the extracted methods (for instance ``pull up method'' for clones in sibling classes).
\end{itemize}

This does not mean that clones outside the ``Can be extracted'' category cannot be refactored using method extraction. However, they may require additional transformations, other refactoring techniques, etc. Because of that, we took those categories out of the scope for our automated refactoring efforts. We further explain opportunities to refactor such clones outside the ``Can be extracted'' category in Section~\ref{sec:discussrefactorability}.

@@

\subsubsection{Spans Part Of A Block}
When applying an ``Abstract Method'' refactoring, we author a new method declaration. The body of this method declaration consists of a set of statements. However, if a single statement is only partially cloned, this might give issues with refactoring. An example of this is shown in Figure~\ref{fig:refactorabilitypartialblock}

@@

\subsubsection{Is Not A Partial Method}
The ``Extract Method'' refactoring can only refactor code in the body of a method. We mapped the contents of clones in Section~\ref{sec:setuplocation}. Here we defined the ``Partial Method'' catagory as \textit{a clone that spans (a part of) the body of a method. The declaration itself is not included.} If the clone does not fall within this category, ``Extract Method'' is likely not to be correct technique for refactoring such clones.

\subsubsection{Top-level Node Is Not A Statement}
A method body can only consist of statements. We cannot put declarations in a method. However, there are possibilities where a clone consists of something else than a statement as top level node. The most prominent example is when a clause is cloned (see Section~\ref{sec:astbackground} for examples of such concepts). For instance, we cannot extract only the catch-clause of a try-statement to a new method. In such a case, we should either extract the complete try-statement, or only the contents of the catch clause.

\subsubsection{Overlaps}
We do not consider clone classes for refactoring that overlap in the clone class itself. That is because such clones may require other techniques to refactor them. We consider analyzing such refactoring methods future work. An example refactoring of clone instances that overlap within their clone class is displayed in Figure~\ref{fig:overlaprefactoring}. In this example one clone instance is found between line \ref{line:overlaprefactoring1} and \ref{line:overlaprefactoring4} and the other between \ref{line:overlaprefactoring2} and \ref{line:overlaprefactoring5}. To refactor such clones, other techniques may be required, like introducing a new for-loop to reduce the duplication.

@@

\subsubsection{Relation}
For each of the identified clone relations (Section~\ref{sec:setuprelation}) we define a method of automatically refactoring clones in these categories. The relation determines the location at which we place the newly created method.
